---
title: "NRRRF_sidestream_data"
author: "Nicholas Chance"
date: "2026-03-20"
output: html_document
---

```{r}
knitr::opts_chunk$set(fig.width = 30, fig.height=10)
```


# library
```{r}
library(reshape2)
library(ggplot2)
library(csv) # for finalized saved csv
library(readxl) # for working documents in excel
```

# Pearson Correlation
# Bring in data
```{r}
IFAS<- as.csv("C:/Users/nicho/OneDrive/IFAS_only_cleaned.csv", 
                row.names = 1, header = TRUE, sep = ",", 
                check.names = TRUE, stringsAsFactors = TRUE)
#remove collection date column
IFAS<-IFAS[,-which(names(IFAS)=="CollectDate.1")]

MBBR<-as.csv("C:/Users/nicho/OneDrive/MBBR_only_cleaned.csv", 
                row.names = 1, header = TRUE, sep = ",", 
                check.names = TRUE, stringsAsFactors = TRUE)
MBBR<-MBBR[,-which(names(MBBR)=="CollectDate.1")]
```

# IFAS
## start corr matrix
```{r}
cormat_IFAS<- round(cor(IFAS, use = "pairwise.complete.obs"),2)
head(cormat_IFAS)
```

```{r}
melt_cormat_IFAS<-melt(cormat_IFAS)
head(melt_cormat_IFAS)
```

## Plot
```{r}
ggplot(data=melt_cormat_IFAS, aes(x=Var1, y=Var2, fill=value))+
  geom_tile()
```

### get lower triangle of corrmatrix
```{r}
get_lower_tri<-function(cormat_IFAS){
  cormat_IFAS[upper.tri(cormat_IFAS)]<-NA
  return(cormat_IFAS)}

lower_tri<-get_lower_tri(cormat_IFAS)
head(lower_tri)
```

```{r}
tri_cormat_ifas<-melt(lower_tri, na.rm=TRUE)

ggplot(data=tri_cormat_ifas, aes(x=Var1,y=Var2, fill=value))+
  geom_tile(color="white")+
  scale_fill_gradient2(low="blue",high="red", mid="white",
                       midpoint=0, limit=c(-1,1), space="Lab",
                       name="Pearson\nCorrelation")+
  theme_minimal()+
  theme(axis.text.x=element_text(angle=90, vjust=1, hjust=1))+
  coord_fixed()
```
## just correlating hzsb to paramters
```{r}
x_vars<- c("ML","New","Preseed")
y_vars<-colnames(IFAS)

IFAS_cor<-round(cor(x=IFAS[,x_vars], y=IFAS[,y_vars], use = "pairwise.complete.obs"),2)
IFAS_cor<-melt(IFAS_cor, na.rm=TRUE)
head(IFAS_cor)
```


```{r}

IFAS_heatmap<-ggplot(data=IFAS_cor, aes(x=Var2,y=Var1, fill=value))+
  geom_tile(color="white")+
  scale_fill_gradient2(low="blue",high="red", mid="white",
                       midpoint=0, limit=c(-1,1), space="Lab",
                       name="Pearson\nCorrelation")+
  theme_minimal()+
  theme(axis.text.x=element_text(angle=45, vjust=1, hjust=1))
```

## add correaltion coefficient to boxes
```{r}
IFAS_heatmap+
  geom_text(aes(Var2, Var1, label=value), color="black", size=3.1)+
  theme(
    axis.title.x=element_blank(),
    axis.title.y=element_blank(),
    panel.grid.major=element_blank(),
    panel.border=element_blank(),
    panel.background=element_blank(),
    axis.ticks=element_blank(),
    legend.justification=c(0.5,0.5),
    legend.direction="horizontal",
    legend.position="top",)+
  guides(fill=guide_colorbar(barwidth=7, barheight=1, title.position="top", title.hjust=0.5))

```

# MBBR

## just correlating hzsb to paramters
```{r}
x_vars<- c("ML","New","Preseed")
y_vars<-colnames(MBBR)

MBBR_cor<-round(cor(x=MBBR[,x_vars], y=MBBR[,y_vars], use = "pairwise.complete.obs"),2)
MBBR_cor<-melt(MBBR_cor, na.rm=TRUE)
head(MBBR_cor)
```

```{r}
MBBR_heatmap<-ggplot(data=MBBR_cor, aes(x=Var2,y=Var1, fill=value))+
  geom_tile(color="white")+
  scale_fill_gradient2(low="blue",high="red", mid="white",
                       midpoint=0, limit=c(-1,1), space="Lab",
                       name="Pearson\nCorrelation")+
  theme_minimal()+
  theme(axis.text.x=element_text(angle=45, vjust=1, hjust=1))
```

## add correaltion coefficient to boxes
```{r}
MBBR_heatmap_pic<-MBBR_heatmap+
  geom_text(aes(Var2, Var1, label=value), color="black", size=3)+
  theme(
    axis.title.x=element_blank(),
    axis.title.y=element_blank(),
    panel.grid.major=element_blank(),
    panel.border=element_blank(),
    panel.background=element_blank(),
    axis.ticks=element_blank(),
    legend.justification=c(0.5,0.5),
    legend.position="top",
    legend.direction="horizontal")+
  guides(fill=guide_colorbar(barwidth=7, barheight=1, title.position="top", title.hjust=0.5))

plot(MBBR_heatmap_pic)

```

## save image
```{r}
ggsave(MBBR_heatmap_pic, filename="MBBR_heatmap.png", height=3, width=12, units= "in", dpi=300)
```


# Testing differences between biomass and DNA normalized correlations

#IFAS biomass pearson
```{r}
IFAS_b<- read_excel("C:/Users/njchance/OneDrive/Anitamox_influent_data_working.xlsx", sheet="IFAS only cleaned")
#remove collection date column
IFAS_b<-IFAS_b[,-which(names(IFAS_b)=="CollectDate...1")]
```

## correlation matrix (pearson)
```{r}
cormat_IFASb<- round(cor(IFAS_b, use = "pairwise.complete.obs"),2)
melt_cormat_IFASb<-melt(cormat_IFASb)
```

## plot
```{r}
ggplot(data=melt_cormat_IFASb, aes(x=Var1, y=Var2, fill=value))+
  geom_tile()
```

```{r}
get_lower_tri<-function(cormat_IFASb){
  cormat_IFASb[upper.tri(cormat_IFASb)]<-NA
  return(cormat_IFASb)}

lower_tri<-get_lower_tri(cormat_IFASb)
head(lower_tri)
```

```{r}
tri_cormat_ifasb<-melt(lower_tri, na.rm=TRUE)

ggplot(data=tri_cormat_ifasb, aes(x=Var1,y=Var2, fill=value))+
  geom_tile(color="white")+
  scale_fill_gradient2(low="blue",high="red", mid="white",
                       midpoint=0, limit=c(-1,1), space="Lab",
                       name="Pearson Correlation")+
  theme_minimal()+
  theme(axis.text.x=element_text(angle=90, vjust=1, hjust=1))+
  coord_fixed()
```
## just correlating hzsb to paramters
```{r}
x_vars<- c("ML hzsB","BF new hzsB")
y_vars<-colnames(IFAS_b)

IFASb_cor<-round(cor(x=IFAS_b[,x_vars], y=IFAS_b[,y_vars], use = "pairwise.complete.obs"),2)
IFASb_cor<-melt(IFASb_cor, na.rm=TRUE)
head(IFASb_cor)
```


```{r}

IFASb_heatmap<-ggplot(data=IFASb_cor, aes(x=Var2,y=Var1, fill=value))+
  geom_tile(color="white")+
  scale_fill_gradient2(low="blue",high="red", mid="white",
                       midpoint=0, limit=c(-1,1), space="Lab",
                       name="Pearson Correlation")+
  theme_minimal()+
  theme(axis.text.x=element_text(angle=45, vjust=1, hjust=1))
```

## add correaltion coefficient to boxes
```{r}
#IFASb_heat_image<-
IFASb_heatmap+
  geom_text(aes(Var2, Var1, label=value), color="black", size=2.5)+
  theme(
    axis.title.x=element_blank(),
    axis.title.y=element_blank(),
    panel.grid.major=element_blank(),
    panel.border=element_blank(),
    panel.background=element_blank(),
    axis.ticks=element_blank(),
    legend.justification=c(0.5,0.5),
    legend.direction="horizontal",
    legend.position="top",)+
  guides(fill=guide_colorbar(barwidth=7, barheight=1, title.position="top", title.hjust=0.5))

```
## Save image
```{r}
ggsave(IFASb_heat_image, filename="IFASb_heatmap.png", height=3, width=12, units= "in", dpi=300)
```

# IFAS DNA pearson
```{r}
IFAS_d<- read_excel("Anitamox_influent_data_working.xlsx", sheet="IFAS only DNA cleaned")
#remove collection date column
IFAS_d<-IFAS_d[,-which(names(IFAS_d)=="CollectDate...13")] # remove both collect dates so that all is numeric
```

## correlation matrix (pearson)
```{r}
cormat_IFASd<- round(cor(IFAS_d, use = "pairwise.complete.obs"),2)
melt_cormat_IFASd<-melt(cormat_IFASd)
```

## plot
```{r}
get_lower_tri<-function(cormat_IFASd){
  cormat_IFASd[upper.tri(cormat_IFASd)]<-NA
  return(cormat_IFASd)}

lower_tri<-get_lower_tri(cormat_IFASd)
head(lower_tri)
```


```{r}
tri_cormat_ifasd<-melt(lower_tri, na.rm=TRUE)

ggplot(data=tri_cormat_ifasd, aes(x=Var1,y=Var2, fill=value))+
  geom_tile(color="white")+
  scale_fill_gradient2(low="blue",high="red", mid="white",
                       midpoint=0, limit=c(-1,1), space="Lab",
                       name="Pearson Correlation")+
  theme_minimal()+
  theme(axis.text.x=element_text(angle=90, vjust=1, hjust=1))+
  coord_fixed()
```

# ML- correlating hzsb to parameters DNA vs BM normalized
```{r}
x_vars<- c("ML hzsB","ML hzsB bm")
y_vars<-colnames(IFAS_d)

IFASd_cor<-round(cor(x=IFAS_d[,x_vars], y=IFAS_d[,y_vars], use = "pairwise.complete.obs"),2)
IFASd_cor<-melt(IFASd_cor, na.rm=TRUE)
head(IFASd_cor)
```

```{r}

IFASd_heatmap<-ggplot(data=IFASd_cor, aes(x=Var2,y=Var1, fill=value))+
  geom_tile(color="white")+
  scale_fill_gradient2(low="blue",high="red", mid="white",
                       midpoint=0, limit=c(-1,1), space="Lab",
                       name="Pearson Correlation")+
  theme_minimal()+
  theme(axis.text.x=element_text(angle=45, vjust=1, hjust=1))
```

## add correaltion coefficient to boxes
```{r}
IFASml_heat_image<-IFASd_heatmap+
  geom_text(aes(Var2, Var1, label=value), color="black", size=2)+
  theme(
    axis.title.x=element_blank(),
    axis.title.y=element_blank(),
    panel.grid.major=element_blank(),
    panel.border=element_blank(),
    panel.background=element_blank(),
    axis.ticks=element_blank(),
    legend.justification=c(0.5,0.5),
    legend.direction="horizontal",
    legend.position="top",)+
  guides(fill=guide_colorbar(barwidth=7, barheight=1, title.position="top", title.hjust=0.5))

```

## save image
```{r}
ggsave(IFASml_heat_image, filename="IFASml_heat_image.png", height=3, width=12, units= "in", dpi=300)
```


# BF new- correlating hzsb to parameters DNA vs BM normalized
```{r}
x_vars<- c("BF new hzsB","BF new hzsB bm")
y_vars<-colnames(IFAS_d)

IFASd_cor<-round(cor(x=IFAS_d[,x_vars], y=IFAS_d[,y_vars], use = "pairwise.complete.obs"),2)
IFASd_cor<-melt(IFASd_cor, na.rm=TRUE)
head(IFASd_cor)
```

```{r}

IFASd_heatmap<-ggplot(data=IFASd_cor, aes(x=Var2,y=Var1, fill=value))+
  geom_tile(color="white")+
  scale_fill_gradient2(low="blue",high="red", mid="white",
                       midpoint=0, limit=c(-1,1), space="Lab",
                       name="Pearson Correlation")+
  theme_minimal()+
  theme(axis.text.x=element_text(angle=45, vjust=1, hjust=1))
```

## add correaltion coefficient to boxes
```{r}
IFASbf_heat_image<-IFASd_heatmap+
  geom_text(aes(Var2, Var1, label=value), color="black", size=2)+
  theme(
    axis.title.x=element_blank(),
    axis.title.y=element_blank(),
    panel.grid.major=element_blank(),
    panel.border=element_blank(),
    panel.background=element_blank(),
    axis.ticks=element_blank(),
    legend.justification=c(0.5,0.5),
    legend.direction="horizontal",
    legend.position="top",)+
  guides(fill=guide_colorbar(barwidth=7, barheight=1, title.position="top", title.hjust=0.5))

```

## save image
```{r}
ggsave(IFASbf_heat_image, filename="IFASbf_heat_image.png", height=3, width=12, units= "in", dpi=300)
```


# MBBR DNA pearson
```{r}
MBBR_d<- read_excel("Anitamox_influent_data_working.xlsx", sheet="MBBR Only DNA cleaned")
#remove collection date column
MBBR_d<-MBBR_d[,-which(names(MBBR_d)=="CollectDate...13")] # remove both collect dates so that all is numeric
```

## correlation matrix (pearson)
```{r}
cormat_MBBRd<- round(cor(MBBR_d, use = "pairwise.complete.obs"),2)
melt_cormat_MBBRd<-melt(cormat_MBBRd)
```

# ML- correlating hzsB and amoA to parameters DNA vs BM normalized
```{r}
x_vars<- c("ML  hzsB","ML hzsB bm", "ML amoA","ML amoA bm")
y_vars<-colnames(MBBR_d)

MBBRd_cor<-round(cor(x=MBBR_d[,x_vars], y=MBBR_d[,y_vars], use = "pairwise.complete.obs"),2)
MBBRd_cor<-melt(MBBRd_cor, na.rm=TRUE)
head(MBBRd_cor)
```


```{r}

MBBRd_heatmap<-ggplot(data=MBBRd_cor, aes(x=Var2,y=Var1, fill=value))+
  geom_tile(color="white")+
  scale_fill_gradient2(low="blue",high="red", mid="white",
                       midpoint=0, limit=c(-1,1), space="Lab",
                       name="Pearson Correlation")+
  theme_minimal()+
  theme(axis.text.x=element_text(angle=45, vjust=1, hjust=1))
```

## add correaltion coefficient to boxes
```{r}
MBBRml_heat_image<-MBBRd_heatmap+
  geom_text(aes(Var2, Var1, label=value), color="black", size=1.8)+
  theme(
    axis.title.x=element_blank(),
    axis.title.y=element_blank(),
    panel.grid.major=element_blank(),
    panel.border=element_blank(),
    panel.background=element_blank(),
    axis.ticks=element_blank(),
    legend.justification=c(0.5,0.5),
    legend.direction="horizontal",
    legend.position="top",)+
  guides(fill=guide_colorbar(barwidth=7, barheight=1, title.position="top", title.hjust=0.5))

```

## save image
```{r}
ggsave(MBBRml_heat_image, filename="MBBRml_heat_image.png", height=3, width=12, units= "in", dpi=300)
```

# BF new- correlating hzsb to parameters DNA vs BM normalized
```{r}
x_vars<- c("BF new hzsB","BF new hzsB bm","BF new amoA","BF new amoA bm")
y_vars<-colnames(MBBR_d)

MBBRd_cor<-round(cor(x=MBBR_d[,x_vars], y=MBBR_d[,y_vars], use = "pairwise.complete.obs"),2)
MBBRd_cor<-melt(MBBRd_cor, na.rm=TRUE)
head(MBBRd_cor)
```


```{r}

MBBRd_heatmap<-ggplot(data=MBBRd_cor, aes(x=Var2,y=Var1, fill=value))+
  geom_tile(color="white")+
  scale_fill_gradient2(low="blue",high="red", mid="white",
                       midpoint=0, limit=c(-1,1), space="Lab",
                       name="Pearson Correlation")+
  theme_minimal()+
  theme(axis.text.x=element_text(angle=45, vjust=1, hjust=1))
```

## add correaltion coefficient to boxes
```{r}
MBBRbf_heat_image<-MBBRd_heatmap+
  geom_text(aes(Var2, Var1, label=value), color="black", size=1.8)+
  theme(
    axis.title.x=element_blank(),
    axis.title.y=element_blank(),
    panel.grid.major=element_blank(),
    panel.border=element_blank(),
    panel.background=element_blank(),
    axis.ticks=element_blank(),
    legend.justification=c(0.5,0.5),
    legend.direction="horizontal",
    legend.position="top",)+
  guides(fill=guide_colorbar(barwidth=7, barheight=1, title.position="top", title.hjust=0.5))

```

## save image
```{r}
ggsave(MBBRbf_heat_image, filename="MBBRbf_heat_image.png", height=3, width=12, units= "in", dpi=300)
```


# Spearman Correlation
## Testing differences between biomass and DNA normalized correlations
#IFAS
```{r}
IFAS_d<- read_excel("Anitamox_influent_data_working.xlsx", sheet="IFAS only DNA cleaned") # despite the name, this sheet has data from both the DNA and biomass normalized ddpcr results

#remove collection date column
IFAS_d<-IFAS_d[,-which(names(IFAS_d)=="CollectDate...13")] # remove both collect dates so that all is numeric
```

## ML- correlating hzsb to parameters DNA vs BM normalized
```{r}
x_vars<- c("ML hzsB","ML hzsB bm")
y_vars<-colnames(IFAS_d)

IFASd_cor<-round(cor(x=IFAS_d[,x_vars], y=IFAS_d[,y_vars], use = "pairwise.complete.obs", method="spearman"),2)
IFASd_cor<-melt(IFASd_cor, na.rm=TRUE)
head(IFASd_cor)
```

```{r}

IFASd_heatmap<-ggplot(data=IFASd_cor, aes(x=Var2,y=Var1, fill=value))+
  geom_tile(color="white")+
  scale_fill_gradient2(low="blue",high="red", mid="white",
                       midpoint=0, limit=c(-1,1), space="Lab",
                       name="Spearman Correlation")+
  theme_minimal()+
  theme(axis.text.x=element_text(angle=45, vjust=1, hjust=1))
```

## add correaltion coefficient to boxes
```{r}
IFASml_heat_image_spear<-IFASd_heatmap+
  geom_text(aes(Var2, Var1, label=value), color="black", size=2)+
  theme(
    axis.title.x=element_blank(),
    axis.title.y=element_blank(),
    panel.grid.major=element_blank(),
    panel.border=element_blank(),
    panel.background=element_blank(),
    axis.ticks=element_blank(),
    legend.justification=c(0.5,0.5),
    legend.direction="horizontal",
    legend.position="top",)+
  guides(fill=guide_colorbar(barwidth=7, barheight=1, title.position="top", title.hjust=0.5))

```

## save image
```{r}
ggsave(IFASml_heat_image_spear, filename="IFASml_heat_image_spear.png", height=3, width=12, units= "in", dpi=300)
```


# BF new- correlating hzsb to parameters DNA vs BM normalized
```{r}
x_vars<- c("BF new hzsB","BF new hzsB bm")
y_vars<-colnames(IFAS_d)

IFASd_cor<-round(cor(x=IFAS_d[,x_vars], y=IFAS_d[,y_vars], use = "pairwise.complete.obs", method="spearman"),2)
IFASd_cor<-melt(IFASd_cor, na.rm=TRUE)
head(IFASd_cor)
```

```{r}

IFASd_heatmap<-ggplot(data=IFASd_cor, aes(x=Var2,y=Var1, fill=value))+
  geom_tile(color="white")+
  scale_fill_gradient2(low="blue",high="red", mid="white",
                       midpoint=0, limit=c(-1,1), space="Lab",
                       name="Spearman Correlation")+
  theme_minimal()+
  theme(axis.text.x=element_text(angle=45, vjust=1, hjust=1))
```

## add correaltion coefficient to boxes
```{r}
IFASbf_heat_image_spear<-IFASd_heatmap+
  geom_text(aes(Var2, Var1, label=value), color="black", size=2)+
  theme(
    axis.title.x=element_blank(),
    axis.title.y=element_blank(),
    panel.grid.major=element_blank(),
    panel.border=element_blank(),
    panel.background=element_blank(),
    axis.ticks=element_blank(),
    legend.justification=c(0.5,0.5),
    legend.direction="horizontal",
    legend.position="top",)+
  guides(fill=guide_colorbar(barwidth=7, barheight=1, title.position="top", title.hjust=0.5))

```

## save image
```{r}
ggsave(IFASbf_heat_image_spear, filename="IFASbf_heat_image_spear.png", height=3, width=12, units= "in", dpi=300)
```
